Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2022-01-10 analysis/paper/paper.Rmd
added a few comments and made a few editorial revisions.

The spring abundance estimate was smaller than fall abundance estimates. Our approach assumed equal capture probabilities between the fall and spring sample efforts. Yet, we suspect that the capture probability in the spring could be lower as higher spring flows reduce Northern Pikeminnow densities biasing our spring abundance estimate low. Therefore, we believe our spring abundance estimate to be conservative.

@@

<!-- Again, some of the following numbers need to be updated after results are formalized -->
Recovery of the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by 26.4 to 145.6 adults, annually. This is between 4.7% and 26.1% of the total mean adult escapement<!--of 558--> reported for the Upper Salmon River MPG in 2017 - 2019 [@Kinzer2020]. These estimates only take into account predation on juvenile Chinook Salmon by Northern Pikeminnow, but presumably, juvenile steelhead and Sockeye Salmon are also prey items in Deadwater Slough, including hatchery releases. Therefore, adult returns of steelhead and Sockeye Salmon are affected as well, including for hathchery populations which provide for recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including multiple ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for emigrating salmonids, including for multiple species and populations, the impact of that benefit could be higher than tributary rehabilitation actions, which typically benefit a single population. Moreover, the deepened, slack water conditions that favor Northern Pikeminnow at Deadwater Slough are a result of the failure of a manmade mining reservoir dam. Taken together, the Deadwater Slough is a favorable candidate for management or restoration actions to benefit local Chinook Salmon populations.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (citation). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local businesses rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require year of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created an almost 2-kilometer section of unnaturally slow and deep river, resulting in favorable conditions for Northern Pikeminnow. We estimate that greater than 10,000 Northern Pikeminnow occupy Deadwater Slough with densities greater than reported elsewhere. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids emigrating through Deadwater Slough including Chinook Salmon i.e., juvenile Chinook Salmon have slower movement rates and little refugia to avoid predation. We estimate that predation from Northern Pikeminnow in Deadwater Slough may reduce adult returns by as much as 146 or 26% or recent returns to the Upper Salmon River MPG. Therefore, predation in Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

>>>>

The spring abundance estimate was smaller than fall abundance estimates. Our approach assumed equal capture probabilities between the fall and spring sample efforts. Yet, we suspect that the capture probability in the spring could be lower as higher spring flows reduce Northern Pikeminnow densities biasing our spring abundance estimate low. Therefore, we believe our spring abundance estimate to be conservative.
<!-- how do higher spring flows reduce pikeminnow densities? If pikeminnow densities are lower, then the total abundance SHOULD be lower. Makes more sense to me to say that higher flows directly lower capture probabilities. -->

@@

<!-- Again, some of the following numbers need to be updated after results are formalized -->
Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by 26.4 to 145.6 adults, annually. This is between 4.7% and 26.1% of the total mean adult escapement<!--of 558--> reported for the Upper Salmon River MPG in 2017 - 2019 [@Kinzer2020]. These estimates only take into account predation on juvenile Chinook Salmon by Northern Pikeminnow, but presumably, juvenile steelhead and Sockeye Salmon are also prey items in Deadwater Slough, including hatchery releases. Therefore, adult returns of steelhead and Sockeye Salmon are affected as well, including for hatchery populations which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including multiple ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the impact of that benefit could be higher than tributary rehabilitation actions which typically benefit a single population. Moreover, the deepened, slack water conditions that favor Northern Pikeminnow at Deadwater Slough are a result of the failure of a manmade mining reservoir dam. Taken together, the Deadwater Slough is a favorable candidate for management or restoration actions to benefit local Chinook Salmon populations.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (*citation*). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local businesses rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require years of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created an almost 2-kilometer section of unnaturally slow and deep river, resulting in favorable conditions for Northern Pikeminnow. We estimate that greater than 10,000 Northern Pikeminnow occupy Deadwater Slough with densities greater than reported elsewhere. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids emigrating through Deadwater Slough including Chinook Salmon i.e., juvenile Chinook Salmon have slower movement rates and little refugia to avoid predation. We estimate that predation from Northern Pikeminnow in Deadwater Slough may reduce adult returns by as much as 146 or 26% of recent returns to the Upper Salmon River MPG. Therefore, predation in Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.
